Location and Scale-Invariant Power Transformations for Transforming Data to

### 2025-04-09 manuscript_rmarkdown/manuscript.Rmd
Update discussion.

In addition to no power transformation and location and scale-invariant power transformations,
conventional and Raymaekers and Rousseeuw's robust adaptation [@Raymaekers2024-zf] were assessed.
For the latter two methods, normalisation before standardisation using was additionally assessed using
the following two methods:

@@

Robust location- and scale-invariant transformations are a suitable replacement for their conventional counterparts.
They demonstrated robustness against outliers and prevent inaccurate transformations and potential numerical issues due to location and scale of the distribution of a feature.
This is particularly relevant for automated data processing, where such issues may go unnoticed.
Compared to non-robust location- and scale-invariant transformations, real-world examples with outliers showed higher residual errors of transformed features. 
Robust transformations seek to minimise residual errors for the central part of the distribution, instead of the entire distribution, including outliers.
The empirical central normality test indicated that robust transformations are better able to achieve central normality in the presence of outliers.
However, in a machine learning experiment of 231 real-world datasets that contained at least one numeric feature, we did not find a meaningful benefit -- nor detriment -- to model performance for location- and scale-invariant power transformations.
One reason may be that numeric features with large location shifts ($|\mu| > 1000.0$) were uncommon.
Of the 4886 numeric features in the 231 datasets, 266 (5%) features in 34 datasets had large location shifts, of which 200 appeared in just 2 datasets.
For the latter two datasets, the transformation method did not show significant difference between groups (Friedman test; $p > 0.05$).
```{r include = FALSE, eval = TRUE}

@@

Location- and scale-invariant transformations are realised by simultaneously optimising three parameters, i.e. transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$.
We derived the log-likelihood function to facilitate optimisation using MLE.
Alternatively, standardisation of a numeric variable (e.g., through subtracting its median value and division by its interquartile range) prior to conventional power transformations may achieve a similar effect in reducing sensitivity to the distribution's location and scale.
While this alternative helps prevent these issues -- provided that normalisation does not lead to negative values for Box-Cox transformation -- location- and scale-invariant transformations seem to provide an overall better transformation to normality (Appendix F).

We assessed several methods for robust power transformation.
Methods that relied on the z-score of the transformed feature or the residual error yielded worse results than the non-robust method.
This is partly due to the initial choice of threshold parameters.
Using different initial values, closer to the upper limits ($\delta_1 = 10.0; \delta_2 = 10.0$), led to a reduced loss.
However, these values effectively correspond to a non-robust transformation, where all instances receive the same weight.
Underperformance of these methods could be explained by their use of transformed features for setting weights.
Consequently, the weights change at each iteration in the MLE optimisation process.
This increases local variance in the log-likelihood function and creates local optima that the optimiser may not handle well.
As a consequence, optimal values for transformation parameter $\lambda$ might differ, which increases the presented loss for optimising threshold parameters.
Methods that relied on the empirical probability did not suffer from this issue, as weights remained fixed during MLE.

We introduced an empirical test for central normality to assess whether distributions deviate from normality in a way that might require closer inspection prior to further processing.
The empirical test for central normality differs from other tests for normality, such as the Shapiro-Wilk test [@Shapiro1965-zd], in that the test statistic is independent of the number of samples.
This makes this test more practical for assessing central normality for larger sample numbers, where other tests may detect inconsequential deviations from normality.

This work has the following limitations.
Firstly, we did observe several numerical stability issues for optimisation criteria other than MLE (Appendix B).
These appear in regions where transformation parameters would lead to very large or small numbers when using conventional power transformations.
For MLE stability issues were not observed.
Secondly, the empirical central normality test is based on simulations instead of statistical theory, and relies on a somewhat arbitrary definition of the central portion of a distribution.
Thus, while the test may asses whether data is sufficiently normally distributed for practical purposes, it should not be used as a strict test for normality.

>>>>

In addition to no power transformation and location- and scale-invariant power transformations,
conventional and Raymaekers and Rousseeuw's robust adaptation [@Raymaekers2024-zf] were assessed.
For the latter two methods, normalisation before standardisation using was additionally assessed using
the following two methods:

@@

Robust location- and scale-invariant transformations are a suitable replacement for their conventional counterparts.
They demonstrated robustness against outliers and prevent inaccurate
transformations and potential numerical issues due to location and scale of the distribution of a feature.
This is particularly relevant for automated data processing, where such issues may go unnoticed.

In simulation, robust location- and scale-invariant transformations ranked best in datasets with outliers,
and ranked highly in datasets without outliers, except for data without outliers that were 
directly derived from strictly normal distributions.
This could be observed in real-world examples, where robust location-and scale-invariant transformations
achieved central normality for when the non-robust variant could not.

However, in a machine learning experiment of 231 real-world datasets that contained at least one numeric feature, 
we did not find a meaningful benefit -- nor detriment -- to model performance for location- and scale-invariant power transformations.
One reason may be that numeric features with large location shifts ($|\mu| > 1000.0$) were uncommon.
Of the 4886 numeric features in the 231 datasets, 266 (5%) features in 34 datasets had large location shifts, of which 200 appeared in just 2 datasets.
For the latter two datasets, the transformation method did not show significant difference between groups (Friedman test; $p > 0.05$).
```{r include = FALSE, eval = TRUE}

@@

Location- and scale-invariant transformations are realised by simultaneously 
optimising three parameters, i.e. transformation parameter $\lambda$, shift parameter $x_0$ and scale parameter $s$.
We derived the log-likelihood function to facilitate optimisation using MLE.
Alternatively, standardisation of a numeric variable 
(e.g., through subtracting its median value and division by its interquartile range) 
prior to conventional power transformations can achieve a similar effect in reducing sensitivity
to the distribution's location and scale, as demonstrated by our findings.
While this alternative helps prevent these issues 
-- provided that normalisation does not lead to negative values for Box-Cox transformations -- 
location- and scale-invariant transformations are able to achieve better transformations
to normality, demonstrated by a lower residual error.

We assessed several weighting methods to achieve robust power transformations.
Methods that relied on the z-score of the transformed feature or the residual error 
yielded worse results than the non-robust method.
Underperformance of these weighting methods could be explained by their reliance on transformed feature values for setting weights.
Consequently, their weights change at each iteration in the MLE optimisation process.
This increases local variance in the log-likelihood function and creates local optima that the optimiser may not handle well.
Methods that relied on the empirical probability did not suffer from this issue, as weights remained fixed during MLE.

We introduced an empirical test for central normality to assess whether sequences
deviate from normality in a way that might require closer inspection prior to further processing.
The empirical test for central normality differs from other tests for normality, 
such as the Shapiro-Wilk test [@Shapiro1965-zd], as it assesses normality of the central portion
of a feature, instead of the entire feature.

This work has the following limitations.
Firstly, we did observe several numerical stability issues for optimisation criteria other than MLE (Appendix B).
These appear in regions where transformation parameters would lead to very large or small numbers when using conventional power transformations.
For MLE stability issues were not observed.
